---
title: "36-460/660 Final Project Code"
author: "Ryan Yu"
date: 04/28/2026
format: pdf
---

# Loading Data

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(brms)
library(ggridges)
library(bayesplot)
library(fuzzyjoin)
```

```{r}
all_screens_df <- read_csv("data/all_screens.csv")
```

## EDA

```{r}
# Screens per player
all_screens_df |> 
  group_by(screener_id) |> 
  summarise(n_screens = n()) |> 
  ggplot() +
  geom_histogram(aes(x=n_screens)) +
  labs(x="Number of Screens", y="Count") +
  theme_minimal()
```

```{r}
# Shot location

ggplot(all_screens_df) +
  geom_point(aes(x=shot_loc_x, y=shot_loc_y), alpha=0.3)
```

```{r}
all_screens_df |> 
  mutate(
    location_x_adj = ifelse(shot_loc_x > 50, 100 - shot_loc_x, shot_loc_x),
    location_x_adj = location_x_adj * 47 / 50, # Rescale [0,50] to [0, 47]
    location_y_adj = shot_loc_y * 50 / 100 # Rescale [0,100] to [0,50]
    ) |> 
  ggplot() +
  geom_point(aes(x=location_y_adj, y=location_x_adj), alpha=0.2) +
  coord_fixed() +
  labs(x="", y="") +
  theme_minimal()
```

```{r}
# Number of ballhandlers per screener
all_screens_df |> 
  group_by(screener_id) |> 
  summarize(n_bh = as.factor(n_distinct(screen_bh))) |> 
  mutate(n_bh = factor(n_bh, levels=sort(as.numeric(levels(n_bh))))) |> 
  ggplot() +
  geom_bar(aes(x=n_bh)) +
  labs(x="Number of Unique Ball Handlers", y="Count", title="Unique Ball Handlers for each Screener") +
  theme_minimal()
```

```{r}
# Number of screeners per ball handler
all_screens_df |> 
  group_by(screen_bh) |> 
  summarize(n_screener = as.factor(n_distinct(screener_id))) |> 
  mutate(n_screener = factor(n_screener, levels=sort(as.numeric(levels(n_screener))))) |> 
  ggplot() +
  geom_bar(aes(x=n_screener)) +
  labs(x="Number of Screen Setters", y="Count", title="Unique Screen Setters for each Ball Handler") +
  theme_minimal()
```

```{r}
all_screens_df |> 
  group_by(screen_type) |> 
  summarize(sum=n())

# condense everything but dho, drag, heavy, rub, an traditional to an "other" category
```

```{r}
all_screens_df |> 
  group_by(screen_location) |> 
  summarize(sum=n())

# Condense corner_3, in_paint, and short_corner to "Other"
```

```{r}
all_screens_df |> 
  group_by(screen_defense) |> 
  summarize(sum=n())

# Collapse get over/go under together, then collapse blitz, ice, jam, none to "Other"
```

```{r}
all_screens_df |> 
  group_by(shot_contest_type) |> 
  summarize(sum=n())

# Collapse kornet into "loose"
```

```{r}
all_screens_df |> 
  group_by(shot_type) |> 
  summarize(sum=n())

# Collapse dunk into layup, then face_up, heave, hook, hook_shot, lob into "Other"
```

```{r}
# collapse fixed effect variables with a low number of observations
all_screens_df <- all_screens_df |> 
  mutate(
    screen_type = ifelse(screen_type %in% c("dho", "rub", "drag", "heavy", "traditional"),
                         screen_type, "other"),
    screen_location = ifelse(screen_location %in% c("corner_3", "in_paint", "short_corner"),
                             "other", screen_location),
    screen_defense = ifelse(str_detect(screen_defense, "get_over"), "get_over",
                            ifelse(str_detect(screen_defense, "go_under"), "go_under",
                                   ifelse(screen_defense %in% c("blitz", "ice", "jam", "none"),
                                          "other", screen_defense
                                          )
                                   )
                            ),
    shot_contest_type = ifelse(shot_contest_type == "kornet", "loose", shot_contest_type),
    shot_type = ifelse(shot_type == "dunk", "layup",
                       ifelse(shot_type %in% c("face_up", "heave", "hook", "hook_shot", "lob"),
                              "other", shot_type))
  )
```

```{r}
# Distribution of plays per screener
all_screens_df |> 
  group_by(screener_id) |> 
  summarize(n_obs = n()) |> 
  ggplot() +
  geom_histogram(aes(x=n_obs)) +
  theme_minimal()

# filter out screeners with fewer than 10 plays
all_screens_df <- all_screens_df |> 
  group_by(screener_id) |> 
  filter(n() >= 10) |> 
  ungroup()

length(unique(all_screens_df$screener_id))
```

```{r}
# Distribution of plays per ball handler
all_screens_df |> 
  group_by(screen_bh) |> 
  summarize(n_obs = n()) |> 
  ggplot() +
  geom_histogram(aes(x=n_obs)) +
  theme_minimal()

# filter out ball handlers with fewer than 10 plays
all_screens_df <- all_screens_df |> 
  group_by(screen_bh) |> 
  filter(n() >= 10) |> 
  ungroup()

length(unique(all_screens_df$screen_bh))
```

```{r}
# Distribution of plays per screener defender
all_screens_df |> 
  group_by(screen_scrd) |> 
  summarize(n_obs = n()) |> 
  ggplot() +
  geom_histogram(aes(x=n_obs)) +
  theme_minimal()

# filter out screener defenders with fewer than 10 plays
all_screens_df <- all_screens_df |> 
  group_by(screen_scrd) |> 
  filter(n() >= 10) |> 
  ungroup()

length(unique(all_screens_df$screen_scrd))
```

```{r}
# Distribution of plays per ball handler defender
all_screens_df |> 
  group_by(screen_bhd) |> 
  summarize(n_obs = n()) |> 
  ggplot() +
  geom_histogram(aes(x=n_obs)) +
  theme_minimal()

# filter out ball handler defenders with fewer than 10 plays
all_screens_df <- all_screens_df |> 
  group_by(screen_bhd) |> 
  filter(n() >= 10) |> 
  ungroup()

length(unique(all_screens_df$screen_scrd))
```

```{r}
# Check multicollinearity
# mm <- model.matrix(make ~ screen_defense + screen_type + screen_location + shot_distance + shot_type + shot_contest_type, data=all_screens_df)
# 
# corrs <- cor(mm[,-1])
```

## Response variable distributions 
     
```{r}
all_screens_df |> 
  ggplot() +
  geom_histogram(aes(x=xpt))
```

```{r}
all_screens_df |> 
  mutate(make = as.factor(make)) |> 
  ggplot() +
  geom_bar(aes(x=make))
```

## Model Fitting

```{r}
# https://www.kaggle.com/datasets/reinerjasin/nba-2k25-player-complete-dataset?select=current_nba_players.csv
nba2k_ratings <- read_csv("data/current_nba_players.csv")
nba2k_ratings <- nba2k_ratings |> 
  select(name, strength, ball_handle, help_defense_iq, perimeter_defense)

player_id_map <- read_csv("data/player_id_map.csv")

nba2k_ratings <- player_id_map |> 
  stringdist_left_join(nba2k_ratings, by="name") |> 
  select(-name.y) |> 
  rename(name = name.x) |> 
  mutate(
    strength = replace_na(strength, mean(strength, na.rm = TRUE)),
    ball_handle = replace_na(ball_handle, mean(ball_handle, na.rm = TRUE)),
    help_defense_iq = replace_na(help_defense_iq, mean(help_defense_iq, na.rm = TRUE)),
    perimeter_defense = replace_na(perimeter_defense, mean(perimeter_defense, na.rm = TRUE))
  )

# manually get rid of duplicate ids caused by fuzzy matching
dup_inds <- c(47, 179, 188, 203, 275, 306, 322, 371, 461, 479)
nba2k_ratings <- nba2k_ratings[-dup_inds, ]

all_screens_df <- all_screens_df |> 
  left_join(nba2k_ratings |> select(id, screener_rating = strength),
            by=join_by(screener_id == id))

all_screens_df <- all_screens_df |> 
  left_join(nba2k_ratings |> select(id, bh_rating = ball_handle),
            by=join_by(screen_bh == id))

all_screens_df <- all_screens_df |> 
  left_join(nba2k_ratings |> select(id, scrd_rating = help_defense_iq),
            by=join_by(screen_scrd == id))

all_screens_df <- all_screens_df |> 
  left_join(nba2k_ratings |> select(id, bhd_rating = perimeter_defense),
            by=join_by(screen_bhd == id))
```

```{r}
all_screens_df <- all_screens_df |> 
  mutate(
    screener_id = as.factor(screener_id),
    screen_bh = as.factor(screen_bh),
    screen_scrd = as.factor(screen_scrd),
    screen_bhd = as.factor(screen_bhd),
    shot_distance = (shot_distance - mean(shot_distance)) / sd(shot_distance)
  )

set.seed(36660)
all_screens_df <- all_screens_df |>
  mutate(fold = sample(c("train", "test"), n(), replace = TRUE, 
                       prob = c(.8, .2)))
train_data <- all_screens_df |>
  filter(fold == "train")
test_data <- all_screens_df |>
  filter(fold == "test")
```

```{r}
screen_make_model <- brm(make ~ 1 + screen_defense + screen_type + screen_location +
                          shot_distance + shot_type + shot_contest_type +
                           screener_rating + bh_rating + scrd_rating + bhd_rating +
                           (1 | screener_id) + (1 | screen_bh) + (1 | screen_scrd) + 
                           (1 | screen_bhd), seed = 36660,
                         family=bernoulli, data=train_data, iter=5000,
                         cores = 12, backend="cmdstanr", file="brms_make_model_v3")
```

```{r}
# screener_only_model <- brm(make ~ screen_defense + screen_type + screen_location +
#                           shot_distance + shot_type + shot_contest_type +
#                            screener_rating + (1 | screener_id),
#                          family=bernoulli, data=train_data, iter=2000,
#                          cores = 12, backend="cmdstanr", file="screener_only_model")
```

# Read in saved data

```{r}
# screen_make_model <- read_rds("brms_make_model.rds")
# make_posterior_sample <- as_tibble(screen_make_model)
```

```{r}
screen_make_model <- read_rds("brms_make_model_v3.rds")
make_posterior_sample <- as_tibble(screen_make_model)
```

# Check distribution of parameters

```{r}
plot(screen_make_model, variable = "^sd_", regex=TRUE)
```

```{r}
pp_check(screen_make_model)
```

```{r}
mcmc_trace(screen_make_model, pars = c("sd_screener_id__Intercept"))
```

# R hat and effective sample size

```{R}
hist(brms::rhat(screen_make_model), main="R hat Distribution", xlab="R hat")
hist(neff_ratio(screen_make_model), main="Effective Sample Size Ratio Distribution", xlab="Effective Sample Size Ratio")
```

# ICC

```{r}
icc <- make_posterior_sample |> 
  select(sd_screener_id__Intercept, sd_screen_bh__Intercept
         #, sd_screen_scrd__Intercept, sd_screen_bhd__Intercept
         ) |> 
  mutate(level2 = sd_screener_id__Intercept^2 + sd_screen_bh__Intercept^2,
         #+ sd_screen_scrd__Intercept^2 + sd_screen_bhd__Intercept^2,
         ICC = level2 / (level2 + pi^2/3))

ggplot(icc) +
  geom_histogram(aes(x=ICC)) +
  labs(y="Frequency", title="Distribution of ICC") +
  theme_minimal()
```

```{r}
library(performance)

icc(screen_make_model)
```

# Variance distributions for different random effects levels

```{r}
sd_posteriors <- make_posterior_sample |> 
  select(contains("sd_")) |> 
  pivot_longer(everything(), 
               names_to = "role", 
               values_to = "estimate") |> 
  # remove annoying text in these parameter names
  mutate(role = ifelse(role == "sd_screener_id__Intercept", "Screener",
                       ifelse(role == "sd_screen_bh__Intercept", "Ball Handler",
                              ifelse(role == "sd_screen_scrd__Intercept", "Screener Defender", "Ball Handler Defender"))))

sd_posteriors |>   
  ggplot(aes(estimate, fill = role)) +
  geom_histogram(alpha = 0.7, bins = 60, show.legend = FALSE) +
  facet_wrap(~ role, nrow = 4) +
  labs(x = "Within-player-role standard deviation estimate", y = "Frequency") + 
  theme_minimal()
```

# Get top screen setters and overperformers

```{r}
screener_posterior <- make_posterior_sample |> 
  select(matches("^r_screener_id"), "b_screener_rating")

screener_posterior_means <- data.frame(
  screener_id = as.numeric(str_sub(colnames(screener_posterior), 15, 19)),
  posterior_mean = colMeans(screener_posterior) 
) |> 
  arrange(desc(posterior_mean))

rownames(screener_posterior_means) <- NULL

screener_posterior_means <- screener_posterior_means |> 
  inner_join(player_id_map, by=join_by(screener_id == id))

screener_top_bottom <- c(head(screener_posterior_means$name, 5),
                   tail(screener_posterior_means$name, 5))

screener_estimates <- screener_posterior |> 
  pivot_longer(cols =-b_screener_rating, names_to = "screener_id", values_to = "estimate") |>  
  mutate(screener_id = as.numeric(str_remove_all(screener_id, "r_screener_id\\[|,Intercept\\]"))) |> # clean up id column
  inner_join(player_id_map, by=join_by(screener_id == id)) |> # filter and also get player names
  mutate(name = fct_reorder(name, estimate, .fun = mean)) # order players by posterior mean

screener_estimates <- screener_estimates |> 
  left_join(nba2k_ratings |> select(id, strength), by=join_by(screener_id == id)) |> 
  mutate(total_effect = b_screener_rating * strength + estimate)

screener_total_means <- screener_estimates |> 
  group_by(name) |> 
  summarize(mean_total_effect = mean(total_effect)) |> 
  arrange(desc(mean_total_effect))

screener_total_top_bottom <- c(head(screener_total_means$name, 5),
                              tail(screener_total_means$name, 5))

screener_estimates |> 
  filter(name %in% screener_total_top_bottom) |> 
  mutate(name = fct_reorder(name, total_effect, .fun = mean)) |>  # order players by posterior mean
  ggplot(aes(total_effect, name)) +
  geom_density_ridges(fill = "lightgray", rel_min_height = 0.01,
                      quantile_lines = TRUE, quantile_fun = mean) +
  labs(x = "Screener total random effect", y = NULL, , title="Top and Bottom 5 Screen Setters") +
  theme_minimal()
```

# Get top ball handlers and overperformers

```{r}
bh_posterior <- make_posterior_sample |> 
  select(matches("^r_screen_bh"), "b_bh_rating")

bh_posterior_means <- data.frame(
  screen_bh = as.numeric(str_sub(colnames(bh_posterior), 13, 17)),
  posterior_mean = colMeans(bh_posterior) 
) |> 
  arrange(desc(posterior_mean))

rownames(bh_posterior_means) <- NULL

bh_posterior_means <- bh_posterior_means |> 
  inner_join(player_id_map, by=join_by(screen_bh == id))

bh_top_bottom <- c(head(bh_posterior_means$name, 5),
                   tail(bh_posterior_means$name, 5))

bh_estimates <- bh_posterior |> 
  pivot_longer(cols=-b_bh_rating, names_to = "screen_bh", values_to = "estimate") |>  
  mutate(screen_bh = as.numeric(str_remove_all(screen_bh, "r_screen_bh\\[|,Intercept\\]"))) |> # clean up id column
  inner_join(player_id_map, by=join_by(screen_bh == id)) |> # filter and also get player names
  mutate(name = fct_reorder(name, estimate, .fun = mean)) # order players by posterior mean

bh_estimates <- bh_estimates |> 
  left_join(nba2k_ratings |> select(id, ball_handle), by=join_by(screen_bh == id)) |> 
  mutate(total_effect = b_bh_rating * ball_handle + estimate)

bh_total_means <- bh_estimates |> 
  group_by(name) |> 
  summarize(mean_total_effect = mean(total_effect)) |> 
  arrange(desc(mean_total_effect))

bh_total_top_bottom <- c(head(bh_total_means$name, 5),
                              tail(bh_total_means$name, 5))

bh_estimates |> 
  filter(name %in% bh_total_top_bottom) |> 
  mutate(name = fct_reorder(name, total_effect, .fun = mean)) |>  # order players by posterior mean
  ggplot(aes(total_effect, name)) +
  geom_density_ridges(fill = "lightgray", rel_min_height = 0.01,
                      quantile_lines = TRUE, quantile_fun = mean) +
  labs(x = "Ball Handler total random effect", y = NULL, , title="Top and Bottom 5 Ball Handlers") +
  theme_minimal()
```

# Get top screener defenders and overperformers

```{r}
scrd_posterior <- make_posterior_sample |> 
  select(matches("^r_screen_scrd"), "b_scrd_rating")

scrd_posterior_means <- data.frame(
  screen_scrd = as.numeric(str_sub(colnames(scrd_posterior), 15, 19)),
  posterior_mean = colMeans(scrd_posterior) 
) |> 
  arrange(desc(posterior_mean))

rownames(scrd_posterior_means) <- NULL

scrd_posterior_means <- scrd_posterior_means |> 
  inner_join(player_id_map, by=join_by(screen_scrd == id))

scrd_top_bottom <- c(head(scrd_posterior_means$name, 5),
                   tail(scrd_posterior_means$name, 5))

scrd_estimates <- scrd_posterior |> 
  pivot_longer(cols=-b_scrd_rating, names_to = "screen_scrd", values_to = "estimate") |>  
  mutate(screen_scrd = as.numeric(str_remove_all(screen_scrd, "r_screen_scrd\\[|,Intercept\\]"))) |> # clean up id column
  inner_join(player_id_map, by=join_by(screen_scrd == id)) |> # filter and also get player names
  mutate(name = fct_reorder(name, estimate, .fun = mean),
         name = fct_rev(name)) # order players by posterior mean

scrd_estimates <- scrd_estimates |> 
  left_join(nba2k_ratings |> select(id, help_defense_iq), by=join_by(screen_scrd == id)) |> 
  mutate(total_effect = b_scrd_rating * help_defense_iq + estimate)

scrd_total_means <- scrd_estimates |> 
  group_by(name) |> 
  summarize(mean_total_effect = mean(total_effect)) |> 
  arrange(desc(mean_total_effect))

scrd_total_top_bottom <- c(head(scrd_total_means$name, 5),
                              tail(scrd_total_means$name, 5))

scrd_estimates |> 
  filter(name %in% scrd_total_top_bottom) |> 
  mutate(name = fct_reorder(name, total_effect, .fun = mean),
         name = fct_rev(name)) |>  # order players by posterior mean
  ggplot(aes(total_effect, name)) +
  geom_density_ridges(fill = "lightgray", rel_min_height = 0.01,
                      quantile_lines = TRUE, quantile_fun = mean) +
  labs(x = "Screener Defender total random effect", y = NULL, , title="Top and Bottom 5 Screener Defenders") +
  theme_minimal()
```

# Get top ball handler defenders

```{r}
bhd_posterior <- make_posterior_sample |> 
  select(matches("^r_screen_bhd"), "b_bhd_rating")

bhd_posterior_means <- data.frame(
  screen_bhd = as.numeric(str_sub(colnames(bhd_posterior), 14, 18)),
  posterior_mean = colMeans(bhd_posterior) 
) |> 
  arrange(desc(posterior_mean))

rownames(bhd_posterior_means) <- NULL

bhd_posterior_means <- bhd_posterior_means |> 
  inner_join(player_id_map, by=join_by(screen_bhd == id))

bhd_top_bottom <- c(head(bhd_posterior_means$name, 5),
                   tail(bhd_posterior_means$name, 5))

bhd_estimates <- bhd_posterior |> 
  pivot_longer(cols=-b_bhd_rating, names_to = "screen_bhd", values_to = "estimate") |>  
  mutate(screen_bhd = as.numeric(str_remove_all(screen_bhd, "r_screen_bhd\\[|,Intercept\\]"))) |> # clean up id column
  inner_join(player_id_map, by=join_by(screen_bhd == id)) |> # filter and also get player names
  mutate(name = fct_reorder(name, estimate, .fun = mean),
         name = fct_rev(name)) # order players by posterior mean

bhd_estimates <- bhd_estimates |> 
  left_join(nba2k_ratings |> select(id, perimeter_defense), by=join_by(screen_bhd == id)) |> 
  mutate(total_effect = b_bhd_rating * perimeter_defense + estimate)

bhd_total_means <- bhd_estimates |> 
  group_by(name) |> 
  summarize(mean_total_effect = mean(total_effect)) |> 
  arrange(desc(mean_total_effect))

bhd_total_top_bottom <- c(head(bhd_total_means$name, 5),
                              tail(bhd_total_means$name, 5))

bhd_estimates |> 
  filter(name %in% bhd_total_top_bottom) |> 
  mutate(name = fct_reorder(name, total_effect, .fun = mean),
         name = fct_rev(name)) |>  # order players by posterior mean
  ggplot(aes(total_effect, name)) +
  geom_density_ridges(fill = "lightgray", rel_min_height = 0.01,
                      quantile_lines = TRUE, quantile_fun = mean) +
  labs(x = "Ball Handler Defender total random effect", y = NULL, , title="Top and Bottom 5 Ball Handler Defenders") +
  theme_minimal()
```

# Accuracy

```{r}
make_logit <- glm(make ~ screen_defense + screen_type + screen_location +
                          shot_distance + shot_type + shot_contest_type,
                  data = train_data,
                  family = binomial)

make_logit_preds <- predict(make_logit, test_data)
make_logit_preds <- ifelse(make_logit_preds > 0.5, 1, 0)
logit_acc <- mean(make_logit_preds == test_data$make)
```

```{r}
test_error_samples <- posterior_epred(screen_make_model, newdata = test_data,
                                      allow_new_levels = TRUE)

acc_basic <- apply(test_error_samples, 1, function(prob) {
  yhat <- ifelse(prob > 0.5, 1, 0)
  mean(yhat == test_data$make)
})

hist(acc_basic, main="Posterior Sample Accuracy Distribution", xlab="Accuracy")
abline(v = logit_acc, col = "red", lty = 2, lwd=2)
legend("topleft", legend=c("Baseline Model Accuracy"), fill=c("red"))
```

```{r}
ifelse(test_error_samples[1, ] > 0.5, 1, 0) |> as.factor() |> plot(main="Posterior Predictive Check")
```
